---
title: "Analysis_Nov15"
author: "Yiming Qian"
date: "11/15/2020"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Fig/',
                      echo=TRUE, warning=FALSE, message=FALSE)
# library(ggplot2)  #for plotting
# library(MASS)     #for random generator function
# library(psych)    #for general functions
# library(nlme)      #for model fitting
# library(reshape)   #for data management
# library(plyr)
library(tidyverse)
# library(GGally)
library(broom)
```

```{r}
setwd("~/R/Cognitive_dev_during_puberty/data")
df3<-read.csv("colorado_dataset.csv", header = TRUE)
options(digits=2)
```
# analysis

### Hidpt1412
```{r}
# model 1-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y12age + y12hidpt, .)) %>% tidy(mod, conf.int = TRUE)

dfHour = df3 %>% group_by(sex) %>%
  do(fitHour = lm(y14hidpt ~ y12age + y12hidpt, data = .))

knitr::kable(tidy(dfHour,fitHour))

# model 1-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y12age +  y12hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = FALSE)

# model 1-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y12age + y12hidpt + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = FALSE)
```

### Crott1412
```{r}
# model 2-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14crott ~ y12age + y12crott, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14crott ~ y12age + y12crott + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14crott ~ y12age + y12crott + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Vocab1412
```{r}
# model 2-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14vocab ~ y12age + y12vocab, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14vocab ~ y12age + y12vocab + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14vocab ~ y12age + y12vocab + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Simil1412
```{r}
# model 2-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14simil ~ y12age + y12simil, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14simil  ~ y12age + y12simil  + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14simil  ~ y12age + y12simil + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Hidpt1614
```{r}
# model 2-1
df3 %>% group_by(sex) %>%
do(mod = lm(y16hidpt ~ y14age + y14hidpt, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-2
df3 %>% group_by(sex) %>%
do(mod = lm(y16hidpt ~ y14age +  y14hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-3
df3 %>% group_by(sex) %>%
do(mod = lm(y16hidpt ~ y14age + y14hidpt + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Crott1614
```{r}
# model 2-1
df3 %>% group_by(sex) %>%
do(mod = lm(y16crott ~ y14age + y14crott, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-2
df3 %>% group_by(sex) %>%
do(mod = lm(y16crott ~ y14age +  y14crott + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 2-3
df3 %>% group_by(sex) %>%
do(mod = lm(y16crott ~ y14age + y14crott + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Hidpt1210
```{r}
# model 3-1
df3 %>% group_by(sex) %>%
do(mod = lm(y12hidpt ~ y10age + y10hidpt, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-2
df3 %>% group_by(sex) %>%
do(mod = lm(y12hidpt ~ y10age + y10hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-3
df3 %>% group_by(sex) %>%
do(mod = lm(y12hidpt ~ y10age + y10hidpt + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Crott1210
```{r}
# model 3-1
df3 %>% group_by(sex) %>%
do(mod = lm(y12crott ~ y10age + y10crott, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-2
df3 %>% group_by(sex) %>%
do(mod = lm(y12crott ~ y10age +  y10crott + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-3
df3 %>% group_by(sex) %>%
do(mod = lm(y12crott ~ y10age + y10crott + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Vocab1210
```{r}
# model 3-1
df3 %>% group_by(sex) %>%
do(mod = lm(y12vocab ~ y10age + y10vocab, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-2
df3 %>% group_by(sex) %>%
do(mod = lm(y12vocab~ y10age +  y10vocab + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-3
df3 %>% group_by(sex) %>%
do(mod = lm(y12vocab ~ y10age + y10vocab + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Simil1210
```{r}
# model 3-1
df3 %>% group_by(sex) %>%
do(mod = lm(y12simil ~ y10age + y10simil, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-2
df3 %>% group_by(sex) %>%
do(mod = lm(y12simil~ y10age +  y10simil + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 3-3
df3 %>% group_by(sex) %>%
do(mod = lm(y12simil ~ y10age + y10simil + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```
### Hidpt1410
```{r}
# model 4-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y10age + y10hidpt, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y10age + y10hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y10age + y10hidpt + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Crott1410
```{r}
# model 4-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14crott ~ y10age + y10crott, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14crott ~ y10age +  y10crott + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14crott ~ y10age + y10crott + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

<<<<<<< HEAD
### Vocab1410
```{r}
# model 4-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14vocab ~ y10age + y10vocab, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14vocab ~ y10age +  y10vocab + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14vocab ~ y10age + y10vocab + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```
### Simil1410
```{r}
# model 4-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14simil ~ y10age + y10simil , .)) %>% tidy(mod, conf.int = TRUE)

# model 4-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14simil  ~ y10age +  y10simil  + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14simil  ~ y10age + y10simil  + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

### Hidpt1216
```{r}
# model 4-1
df3 %>% group_by(sex) %>%
do(mod = lm(y16hidpt ~ y12age + y12hidpt, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-2
df3 %>% group_by(sex) %>%
do(mod = lm(y16hidpt ~ y12age + y12hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

# model 4-3
df3 %>% group_by(sex) %>%
do(mod = lm(y16hidpt ~ y12age + y12hidpt + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = TRUE)
```

## replicate
### 1412
```{r}
# model 1-2
df3 %>% group_by(sex,replicate) %>%
do(mod = lm(y14hidpt ~ y12age +  y12hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

df3 %>% group_by(sex,replicate) %>%
do(mod = lm(y14crott ~ y12age +  y12crott + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

df3 %>% group_by(sex,replicate) %>%
do(mod = lm(y14simil ~ y12age +  y12simil + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)

df3 %>% group_by(sex,replicate) %>%
do(mod = lm(y14vocab ~ y12age +  y12vocab + time_log_excl, .)) %>% tidy(mod, conf.int = TRUE)
```
### 1014
```{r}
df3 %>% group_by(sex,replicate) %>%
do(mod = lm(y14hidpt ~ y10age +  y10hidpt + time_log_excl , .)) %>% tidy(mod, conf.int = TRUE)
```

### Hidpt1412
```{r}
df4<-df3%>%
  filter(y12age >=12)

# model 1-1
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y12age + y12hidpt, .)) %>% tidy(mod, conf.int = TRUE)

dfHour = df3 %>% group_by(sex) %>%
  do(fitHour = lm(y14hidpt ~ y12age + y12hidpt, data = .))

knitr::kable(tidy(dfHour,fitHour))

# model 1-2
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y12age +  y12hidpt + time_log_excl, .)) %>% tidy(mod, conf.int = FALSE)

# model 1-3
df3 %>% group_by(sex) %>%
do(mod = lm(y14hidpt ~ y12age + y12hidpt + time_log_excl +  time_log_sq, .)) %>% tidy(mod, conf.int = FALSE)
=======
```{r}
df3 %>% ggplot(aes(x=time_log_excl, y=y14hidpt, group=sex, color=sex)) + 
  geom_point(alpha=0.4) +
  geom_smooth() +
  labs(x="Puberty timing", y="Hidden pattern at 14")+
  #theme with white background
  theme_classic() 

df3 %>% ggplot(aes(x=time_log_excl, y=hidpt1412, group=sex, color=sex)) + 
  geom_point(alpha=0.4) +
 # facet_grid(.~age) +
  geom_smooth() +
  labs(x="Puberty timing", y="Hidden pattern at 14")+
  #theme with white background
  theme_classic() 
>>>>>>> da0ee75f9537e2313b37415107740a116fa491c2
```