---
title: "Combined_data_visualization_modeling"
author: "Yiming Qian"
date: "6/28/2020"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Fig/',
                      echo=TRUE, warning=FALSE, message=FALSE)
library(ggplot2)  #for plotting
library(MASS)     #for random generator function
library(psych)    #for general functions
library(nlme)      #for model fitting
library(reshape)   #for data management
library(dplyr)
library(plot3D)
library(plyr)
library(tidyverse)
library(GGally)
library(broom)
library(car)
library(lavaan)
```
Step 1: combine the puberty data and cognitive tests data
```{r join}
raw_cogni_long$id <- as.integer(raw_cogni_long$id)
names(raw_cogni_long)[names(raw_cogni_long) == "age"] <- "age_cog"
names(raw_PDS_long)[names(raw_PDS_long) == "age"] <- "age_pds"
a_long <- full_join(raw_PDS_long[-c(6:12)],raw_cogni_long[,-c(1,5)],by=c("id","occasion"))
subset(a_long,round(age_pds,6)!=round(age_cog,6))
a_long$sex<-factor(a_long$sex, levels=c(0,1),labels=c("Girls","Boys"))
a_long$occasion<-factor(a_long$occasion)

raw_PDS_final<- raw_PDS_long[,-c(5:12)] %>%
  pivot_wider(names_from = occasion,values_from = pds)
names(raw_PDS_final)[names(raw_PDS_final) == "y9"] <- "y9pds"
names(raw_PDS_final)[names(raw_PDS_final) == "y10"] <- "y10pds"
names(raw_PDS_final)[names(raw_PDS_final) == "y11"] <- "y11pds"
names(raw_PDS_final)[names(raw_PDS_final) == "y12"] <- "y12pds"
names(raw_PDS_final)[names(raw_PDS_final) == "y13"] <- "y13pds"
names(raw_PDS_final)[names(raw_PDS_final) == "y14"] <- "y14pds"
names(raw_PDS_final)[names(raw_PDS_final) == "y15"] <- "y15pds"
a <- full_join(raw_PDS_final,raw_cogni[,-c(1,4)],by=c("id"))
a$sex<-factor(a$sex, levels=c(0,1),labels=c("Girls","Boys"))
```

Step 2: check the correlation between puberty and spatial ability by year and sex
```{r}
ggpairs(a_long[,6:7],mapping=ggplot2::aes(color=factor(a_long$sex)))+theme_classic()

test_long<-a_long %>% 
  filter(occasion %in% c("y9","y10","y12","y14"))

tgc1 <- Rmisc::summarySE(test_long, measurevar="hidpt", groupvars=c("pds","sex"), na.rm=TRUE)
tgc1 %>% ggplot(aes(x=pds, y=hidpt, colour=sex)) + 
    geom_errorbar(aes(ymin=hidpt-se, ymax=hidpt+se), width=.1) +
    geom_line() +
    geom_point()+
  #  geom_ribbon(aes(ymin=hidpt-se, ymax=hidpt+se))+
  labs(x="Puberty Stages", y="Hidden Pattern")+  #theme with white background
  theme_classic() 

test_long %>% ggplot(aes(x=pds, y=hidpt, group=sex, color=sex)) + 
  geom_point(alpha=0.3) +
  facet_grid(.~occasion) +
  geom_smooth() +
  labs(x="Puberty Stages", y="Hidden pattern")+
  #theme with white background
  theme_classic() 

tgc1 <- Rmisc::summarySE(test_long, measurevar="hidpt", groupvars=c("pds","sex","occasion"), na.rm=TRUE)
tgc1 %>% ggplot(aes(x=pds, y=hidpt, colour=sex)) + 
    geom_errorbar(aes(ymin=hidpt-se, ymax=hidpt+se), width=.1) +
    geom_line() +
    geom_point()+
  facet_grid(.~occasion) +
  labs(x="Puberty Stages", y="Hidden Pattern")+  #theme with white background
  theme_classic() 
```
```{r}
ggpairs(a_long[,c(6,8)],mapping=ggplot2::aes(color=factor(a_long$sex)))+theme_classic()

tgc1 <- Rmisc::summarySE(test_long, measurevar="crott", groupvars=c("pds","sex"), na.rm=TRUE)
tgc1 %>% ggplot(aes(x=pds, y=crott, colour=sex)) + 
    geom_errorbar(aes(ymin=crott-se, ymax=crott+se), width=.1) +
    geom_line() +
    geom_point()+
  labs(x="Puberty Stages", y="crott")+  #theme with white background
  theme_classic() 

test_long %>% ggplot(aes(x=pds, y=crott, group=sex, color=sex)) + 
  geom_point(alpha=0.3) +
  facet_grid(.~occasion) +
  geom_smooth() +
  labs(x="Puberty Stages", y="Crott")+
  #theme with white background
  theme_classic() 
```
```{r 3D-plot}
# # scatter plot
# library(plot3D)
# x<-subset(a_long, id<=1500)$age_cog
# y<-subset(a_long, id<=1500)$pds
# z<-subset(a_long, id<=1500)$hidpt
# scatter3D(x,y,z,phi=0, bty="g", type="b", ticktype="detailed", pch =20,  xlab="age",ylab="pds", zlab="Hidden Pattern", xlim(9,18),ylim(1,4),clab=c("hidden pattern","score"))

library(car)
if(interactive() && require(rgl) && require(mgcv)){
    scatter3d(hidpt ~ age_cog + pds, data=a_long, id=list(n=3))
    Sys.sleep(5) # wait 5 seconds
    scatter3d(hidpt ~ age_cog + pds | sex, data=a_long)
    Sys.sleep(5)
    scatter3d(hidpt ~ age_cog + pds | sex, surface=FALSE,
    	ellipsoid=TRUE, revolutions=3, data=a_long)
    Sys.sleep(5)
}
# group by sex
scatter3d(a_long$age_cog,a_long$pds,a_long$hidpt, group=a_long$sex,  surface=FALSE,
    	ellipsoid=TRUE,  xlab="age",ylab="pds", zlab="Hidden Pattern")
# group by quantile of hidpt at age 9 
scatter3d(a_long$age_cog,a_long$pds,a_long$hidpt, group=a_long$q_hidpt, grid=FALSE, xlab="age",ylab="pds", zlab="Hidden Pattern")
# male
a_long.m<-a_long %>% filter(sex==1)
scatter3d(a_long.m$age_cog,a_long.m$pds,a_long.m$hidpt, group=a_long.m$q_hidpt, surface=FALSE,
    	ellipsoid=TRUE, xlab="age",ylab="pds", zlab="Hidden Pattern")
# group by occasion
test_long$occasion<-factor(test_long$occasion)
scatter3d(test_long$age_cog,test_long$pds,test_long$hidpt, group=test_long$occasion, surface=FALSE,
    	ellipsoid=TRUE, xlab="age",ylab="pds", zlab="Hidden Pattern")

# library(plotly)
# # it does not work, because points are connected according to `x`(?) not id
# plot_ly(subset(a_long, id<=1500), x = ~age_cog, y = ~pds, z = ~hidpt, type = 'scatter3d', mode = 'lines',
#         opacity = 1, line = list(width = 6, color = ~id, reverscale = FALSE))

```
# Step 3: Model Building, Estimation, and Assessment of Fit to Data
The puberty development follow the logistic regression (see Beltz, Corley, Bricker , Wadsworth, & Berenbaum, 2014, which use the dataset from the same project). However, the repeated measures of spatial ability scores were standardized by gender and age. It provides challenges to investigate the association between the trajectory of puberty development and trajectory of spatial ability development.

The question of interest was whether change in puberty (linear, quadratic change, or logistic change) was associated with a similar pattern of change in cognitive ability. (that is, whether change in puberty status relative to average group change is related to cognitive ability relative to the group?)

(a) I have tried dyadic APIM:
```{r}
a_melt <- a[,c("id","sex", "y9pds","y10pds","y12pds","y14pds","y9hidpt","y10hidpt","y12hidpt","y14hidpt")] %>%
  pivot_longer(cols=c("y9pds","y10pds","y12pds","y14pds","y9hidpt","y10hidpt","y12hidpt","y14hidpt"), 
               names_to=c("age","value"), 
               names_pattern = "(y[0-9]{1,2})(\\w{3,5})",
               values_to="key",
               names_ptype=list(
                 age=factor(
                   levels=c("y9", "y10","y12", "y14"),
                   ordered=TRUE))) 

a_melt$age_baseline[a_melt$age == "y9"] <- 0
a_melt$age_baseline[a_melt$age == "y10"] <- 1
a_melt$age_baseline[a_melt$age == "y12"] <- 3
a_melt$age_baseline[a_melt$age == "y14"] <- 5
a_melt$puberty_on <- ifelse(a_melt$value == "pds", 1, 0)
a_melt$hidpt_on <- ifelse(a_melt$value == "hidpt", 1, 0)
a_melt$key<-as.numeric(a_melt$key)
#  a_melt$key[a_melt$key=="NaN"] <- NA
apim_full <- lme(fixed= key ~  -1 +              #turns OFF intercept 
                                 puberty_on +           #pds intercept
                                 puberty_on:age_baseline +
                                 hidpt_on +        # spatial ability intercept
                                 hidpt_on:age_baseline,
                random = ~ -1 +puberty_on + hidpt_on| id, 
               weights=varIdent(form = ~1 | puberty_on), # this invokes separate sigma^{2}_{e} for each gender
                  corr=corAR1(form = ~1 | id/sex/age_baseline), 
               data = a_melt,
               control=list(maxIter=10000),
               na.action = na.exclude)

summary(apim_full)
# not significant
```

PDS as effect on verbal and spatial ability
```{r}
a_melt1 <- a[,c("id","sex", "y9pds","y10pds","y12pds","y14pds","y9hidpt","y10hidpt","y12hidpt","y14hidpt","y9crott","y10crott","y12crott","y14crott")] %>%
   pivot_longer(cols=y9pds:y14crott, 
               names_to=c("age","value"), 
               names_pattern = "(y[0-9]{1,2})(\\w{3,5})",
               values_to="key",
               names_ptype=list(
                 age=factor(
                   levels=c("y9", "y10","y12", "y14"),
                   ordered=TRUE)))

  
  pivot_longer(cols=c("y9crott","y10crott","y12crott","y14crott","y9hidpt","y10hidpt","y12hidpt","y14hidpt"), 
               names_to=c("age","value"), 
               names_pattern = "(y[0-9]{1,2})(\\w{3,5})",
               values_to="key",
               names_ptype=list(
                 age=factor(
                   levels=c("y9", "y10","y12", "y14"),
                   ordered=TRUE))) %>%
 
a_melt1$crott_on <- ifelse(a_melt1$value == "crott", 1, 0)
a_melt1$hidpt_on <- ifelse(a_melt1$value == "hidpt", 1, 0)
a_melt$key<-as.numeric(a_melt$key)
#  a_melt$key[a_melt$key=="NaN"] <- NA
apim_full1 <- lme(fixed= key ~  -1 +              #turns OFF intercept 
                                 crott_on +           #pds intercept
                                 crott_on:pdss +
                                 hidpt_on +        # spatial ability intercept
                                 hidpt_on:pdss,
                random = ~ -1 +crott_on + hidpt_on| id, 
               weights=varIdent(form = ~1 | crott_on), # this invokes separate sigma^{2}_{e} for each gender
                  corr=corAR1(form = ~1 | id/sex/pdss), 
               data = a_melt1,
               control=list(maxIter=10000),
               na.action = na.exclude)

summary(apim_full1)
# not significant
```

(b) SEM to check the correlation of linear slope and quandratic slope

linear model
```{r}
library(lavaan)
Model1<-'L1=~1*y9pds+1*y10pds+1*y11pds+1*y12pds+1*y13pds+1*y14pds+1*y15pds
S1=~0*y9pds+1*y10pds+2*y11pds+3*y12pds+4*y13pds+5*y14pds+6*y15pds
L2=~1*y9hidpt+1*y10hidpt+1*y12hidpt+1*y14hidpt+1*y16hidpt
S2=~0*y9hidpt+1*y10hidpt+2*y12hidpt+3*y14hidpt+4*y16hidpt
L1~~S1
L2~~0*S2
L1~~0*L2
S1~~S2    # completely correlated
L2~~S1
L1~~0*S2
y9pds~0*1
y10pds~0*1
y11pds~0*1
y12pds~0*1
y13pds~0*1
y14pds~0*1
y15pds~0*1
y9hidpt~0*1
y10hidpt~0*1
y12hidpt~0*1
y14hidpt~0*1
y16hidpt~0*1
L1~1
L2~1
S1~1
S2~1'
cfa.bi1<-cfa(Model1, data=a,meanstructure=T)
summary(cfa.bi1,fit.measures=T)
# all the variable is close to 1, so you can consider to set them as the same
inspect(cfa.bi1,'est')
MI<-modificationIndex(mi>5)
MI
```
```{r}
Model2.1<-'L=~1*y9hidpt+1*y10hidpt+1*y12hidpt+1*y14hidpt+1*y16hidpt
S=~-2*y9hidpt+-1*y10hidpt+0*y12hidpt+1*y14hidpt+2*y16hidpt
y10hidpt~y9hidpt
y12hidpt~y10hidpt
y14hidpt~y12hidpt
y16hidpt~y14hidpt'
cfa2.1<-cfa(Model2.1,data=a)
summary(cfa2.1,fit.measures=T)
Model2.2<-'L=~1*y9hidpt+1*y10hidpt+1*y12hidpt+1*y14hidpt+1*y16hidpt
S=~-2*y9hidpt+-1*y10hidpt+0*y12hidpt+1*y14hidpt+2*y16hidpt
Q=~4*y9hidpt+1*y10hidpt+0*y12hidpt+1*y14hidpt+4*y16hidpt
y10hidpt~0*y9hidpt
y12hidpt~0*y10hidpt
y14hidpt~y12hidpt
y16hidpt~y14hidpt'
cfa2.2<-cfa(Model2.2,data=a)
summary(cfa2.2,fit.measures=T)  # better
anova(cfa2.1,cfa2.2)

Model3.1<-'L=~1*y9pds+1*y10pds+1*y11pds+1*y12pds+1*y13pds+1*y14pds+1*y15pds
S=~-3*y9pds+-2*y10pds+-1*y11pds+0*y12pds+1*y13pds+2*y14pds+3*y15pds
y10pds~y9pds
y11pds~y10pds
y12pds~y11pds
y13pds~y12pds
y14pds~y13pds
y15pds~0*y14pds'
cfa3.1<-cfa(Model3.1,data=a)
summary(cfa3.1,fit.measures=T)
Model3.2<-'L=~1*y9pds+1*y10pds+1*y11pds+1*y12pds+1*y13pds+1*y14pds+1*y15pds
S=~-3*y9pds+-2*y10pds+-1*y11pds+0*y12pds+1*y13pds+2*y14pds+3*y15pds
Q=~9*y9pds+4*y10pds+1*y11pds+0*y12pds+1*y13pds+4*y14pds+9*y15pds
y10pds~0*y9pds
y11pds~y10pds
y12pds~y11pds
y13pds~y12pds
y14pds~y13pds
y15pds~0*y14pds'
cfa3.2<-cfa(Model3.2,data=a)
summary(cfa3.2,fit.measures=T)
anova(cfa3.1,cfa3.2)  # better
```

```{r}
# Bivariate standard ALT with covariable
Model4<-'S11=~1*y9pds
S21=~1*y10pds
S31=~1*y11pds
S41=~1*y12pds
S51=~1*y13pds
S61=~1*y14pds
S71=~1*y15pds
S12=~1*y9hidpt
S22=~1*y10hidpt
S32=~1*y12hidpt
S42=~1*y14hidpt
S52=~1*y16hidpt
L1=~1*S11+1*S21+1*S31+1*S41+1*S51+1*S61+1*S71
S1=~-3*S11+-2*S21+-1*S31+0*S41+1*S51+2*S61+3*S71
Q1=~9*S11+4*S21+1*S31+0*S41+1*S51+4*S61+9*S71
L2=~1*S12+1*S22+1*S32+1*S42+1*S52
S2=~-2*S12+-1*S22+0*S32+1*S42+2*S52
Q2=~4*S12+1*S22+0*S32+1*S42+3*S52
S21~0*S11
S31~S21
S41~S31
S51~S41
S61~S51
S71~S61
S22~S12
S32~S22
S42~S32
S52~S42
L1~~L2
L1~~S2
L1~~Q2
S1~~S2
S1~~Q1
L2~~S1
L2~~Q1
S2~~Q1
Q1~~Q2
y9pds~0*1
y10pds~0*1
y11pds~0*1
y12pds~0*1
y13pds~0*1
y14pds~0*1
y15pds~0*1
y9hidpt~0*1
y10hidpt~0*1
y12hidpt~0*1
y14hidpt~0*1
y16hidpt~0*1
L1~1
S1~1
Q1~1
S11~1
S12~1
S11~~S12
S71~~0*S52'
cfa4<-cfa(Model4, data=a)
summary(cfa4, fit.measures=T)
```

(c) growth model

Run univariate model with a time-varying covariate in multilevel framework
The time-varying covariate is called “spring”.
```{r}
hidpt.tvc.nlme<-nlme(hidpt~(beta_1+d_1i)+(beta_2+d_2i)*age_cog+b3*pds,
                      data      = a_long,
                      fixed     = beta_1+beta_2+b3~1,
                      random    = d_1i+d_2i~1,
                      group     = ~id,
                      start     = c(35, 4, 0.1),
                      na.action = na.omit,
                      control = lmeControl(maxIter = 1000))

summary(hidpt.tvc.nlme)
```
Jenss-Bayley Growth Model.
```{r}
hidpt.jb.nlme <- nlme(hidpt~b_1i + b_2i * pds + b_3i * (exp(gamma*pds)-1),
                      data = a_long,
                      fixed = b_1i + b_2i + b_3i + gamma ~ 1,
                      random = b_1i + b_2i + b_3i ~ 1,
                      groups =~ id,
                      start = c(50, 10, 30, -2),
                      na.action = na.exclude)
summary(hidpt.jb.nlme)
```

```{r}
cdlogistic.y15<-nlme(hidpt ~ -8 + 188*(1/(1+exp((gamma_00 + gamma_01*pds + u_0i)*((gamma_10 + gamma_11*pds + u_1i)-age_cog)))),
                    fixed = gamma_00 + gamma_01 + gamma_10 + gamma_11 ~ 1,
                    random = u_0i + u_1i  ~ 1,
                    group = ~ id,
                    start = c(gamma_00= 0.5, gamma_01 = 0.1, gamma_10 = 14, gamma_11=0.1),  #important to get these in the ballpark
                    data = a_long,
                    na.action = "na.exclude",
                    control = lmeControl(maxIter = 200))

summary(cdlogistic.y15)
VarCorr(cdlogistic.y15)
```

Bilinear Spline Growth Model.
```{r}
hidpt.spline.nlme <- nlme(hidpt ~ b_1i + b_2i*(pmin(0,pds-gamma)) + b_3i*(pmax(0,pds-gamma)),
                      data = a_long,
                      fixed = b_1i + b_2i + b_3i + gamma ~ 1,
                      random = b_1i + b_2i + b_3i ~ 1,
                      groups =~ id,
                      start = c(60, 5, 2, 2.5),
                      na.action = na.omit)
summary(hidpt.spline.nlme) 
```