---
title: "challenge_note"
author: "yiming"
date: "April 16, 2020"
output: html_document
---

---
title: "survivalModel"
author: "Yiming Qian"
date: "April 15, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
library(ggplot2)   #for plotting
library(nlme)      #for model fitting
library(psych)     #for general functions
library(reshape)   #for data management
library(dplyr)
library(tidyverse)
library(survival)
library(GGally)
```

# Step 1: Read the data
# prepare the data
```{r}
#set filepath for data file
setwd("C:/Users/yimin/Box Sync/course/HDFS523/final_project")

df_wide <- read.csv("C:/Users/yimin/Box Sync/course/HDFS523/Assignment/Assignment6/df_wide.csv", header=TRUE)
# all.equal(df_wide$y9age/12,df_wide$age9)
# df_wide$y9age/12 == df_wide$age9
# all.equal(df_wide$y10age/12,df_wide$age10)
# df_wide$y10age/12 == df_wide$age10
# all.equal(df_wide$y12age,df_wide$age12)
# all.equal(df_wide$y14age,df_wide$age14)
#Creating a list of variables of interest, only get the pds scores and spatial ability scores which were measured at the same occasions
varnames <- c("project","sex","racecat","hispanic","id","y7space","y9space","y10space","y12space","y14space","y16space","y7verbal","y9verbal","y10verbal","y12verbal","y14verbal","y16verbal","pdss9","pdss10","pdss11","pdss12","pdss13","pdss14","pdss15","time_log_excl")

#sub-setting the columns of interest
a <- df_wide[ ,varnames]
a$sex<-factor(a$sex)
a$racecat<-factor(a$racecat)
a<- a %>% plyr::rename(c("pdss9"="y9pds","pdss10"="y10pds","pdss11"="y11pds","pdss12"="y12pds","pdss13"="y13pds","pdss14"="y14pds","pdss15"="y15pds"))
str(a)
```

## get the long data
```{r}
a_long <- read.csv("C:/Users/yimin/Box Sync/course/HDFS523/Assignment/Assignment9/along.csv", header=TRUE)
a_long$sex<-factor(a_long$sex)
a_long$racecat<-factor(a_long$racecat)
str(a_long)
head(a_long)
```

# Step 2: Data Description/Visualization
## check puberty data 
```{r}
describe(a_long$pds)
describe(a[,c("y9pds","y10pds","y11pds","y12pds","y13pds","y14pds","y15pds")])

pairs.panels(a[,c("y9pds","y10pds","y11pds","y12pds","y13pds","y14pds","y15pds")])

qplot(x=factor(age), y=pds, data=a_long, geom="boxplot", ylab="puberty Status", xlab="Age")

## the distribution of puberty by age
ggplot(data=a_long, aes(x=pds)) + 
  geom_density(aes(group=age, colour=age, fill=age), alpha=0.3) +
  guides(colour=FALSE,
         fill=guide_legend(title="Grade")) +
  labs(x="Puberty Stages", y="Density")+
  ggtitle("Puberty Stages by Grades") +
  #theme with white background
  theme_classic() + 
  #increase font size of axis and point labels
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.position = "right")

## the distribution of puberty by age and sex
ggplot(data=a_long, aes(x=pds)) + 
  geom_density(aes(group=age, colour=age, fill=age), alpha=0.3) +
  guides(colour=FALSE,
         fill=guide_legend(title="Grade")) +
  labs(x="Puberty Stages", y="Density")+
  ggtitle("Puberty Stages by Grades") +
  #theme with white background
  theme_classic() + 
  facet_grid(.~sex)
  #increase font size of axis and point labels
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.position = "right")
```
## check spatial data 
```{r}
describe(a_long$spatial) 

describeBy(a_long$spatial,a_long$sex) 
describeBy(a[,6:11], a$sex)
## the mean and sd of y16space are significantly different from the others
# the mean spatial ability scores for males are 0.71, sd is 1.11

# Correlations plot
pairs.panels(a[,c("y7space","y9space","y10space","y12space","y14space","y16space")])

## the distribution of puberty by age and sex
ggplot(data=a_long, aes(x=spatial)) + 
  geom_density(aes(group=age, colour=age, fill=age), alpha=0.3) +
  guides(colour=FALSE,
         fill=guide_legend(title="Age")) +
  labs(x="Spatial Ability", y="Density")+
  ggtitle("Spatial Ability by Age") +
  #theme with white background
  theme_classic() + 
  facet_grid(.~sex)
  #increase font size of axis and point labels
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.position = "right")
```
## check verbal data 
```{r}
describe(a_long$verbal) 

describeBy(a_long$verbal,a_long$sex) 
describeBy(a[,12:17], a$sex)
```
## check the correlation between puberty and spatial ability by year and sex
```{r}
# check correlation of puberty status and spatial ability scores
# ggpairs(a[,6:18],mapping=ggplot2::aes(color=a$sex))+theme_classic()
ggpairs(a_long[,7:8],mapping=ggplot2::aes(color=a_long$sex))+theme_classic()

tgc1 <- Rmisc::summarySE(a_long, measurevar="spatial", groupvars=c("pds","sex"), na.rm=TRUE)
tgc1 %>% ggplot(aes(x=pds, y=spatial, colour=sex)) + 
    geom_errorbar(aes(ymin=spatial-se, ymax=spatial+se), width=.1) +
    geom_line() +
    geom_point()+
  labs(x="Puberty Stages", y="Spatial ability")+  #theme with white background
  theme_classic() 

a_long %>% ggplot(aes(x=pds, y=spatial, group=sex, color=sex)) + 
  geom_point(alpha=0.3) +
  facet_grid(.~age) +
  geom_smooth() +
  labs(x="Puberty Stages", y="Spatial ability")+
  #theme with white background
  theme_classic() 

tgc <- Rmisc::summarySE(a_long, measurevar="spatial", groupvars=c("pds","sex","age"), na.rm=TRUE)
tgc %>% ggplot(aes(x=pds, y=spatial, colour=sex)) + 
    geom_errorbar(aes(ymin=spatial-se, ymax=spatial+se), width=.1) +
    geom_line() +
    geom_point()+
  facet_grid(.~age) +
  labs(x="Puberty Stages", y="Spatial ability")+  #theme with white background
  theme_classic() 
```

# Step 3: Model Building, Estimation, and Assessment of Fit to Data
The puberty development follow the logistic regression (see Beltz, Corley, Bricker , Wadsworth, & Berenbaum, 2014, which use the dataset from the same project). However, the repeated measures of spatial ability scores were standardized by gender and age. It provides challenges to investigate the association between the trajectory of puberty development and trajectory of spatial ability development.

The question of interest was whether change in puberty (linear, quadratic change, or logistic change) was associated with a similar pattern of change in cognitive ability. (that is, whether change in puberty status relative to average group change is related to cognitive ability relative to the group?)

(a) I have tried dyadic APIM:
```{r, include=FALSE}
# for my own pc
df_wide <- read.csv("C:/Users/yimin/Box Sync/course/HDFS523/Assignment/Assignment6/df_wide.csv", header=TRUE)
#Creating a list of variables of interest, only get the pds scores and spatial ability scores which were measured at the same occasions
varnames <- c("project","sex","racecat","hispanic","id","y9space","y10space","y12space","y14space","pdss9","pdss10","pdss12","pdss14","time_log_excl")
#sub-setting the columns of interest
a <- df_wide[ ,varnames]
a$sex<-factor(a$sex)
a$racecat<-factor(a$racecat)
a<- a %>% rename(y9pds=pdss9,y10pds=pdss10,y12pds=pdss12,y14pds=pdss14)
a_long <- a %>% 
  pivot_longer(cols=y9space:y14pds, 
               names_to=c("age","group"), 
               names_pattern = "(y[0-9]{1,2})(\\w)",
               values_to="key",
               names_ptype=list(
                 age=factor(
                   levels=c("y9", "y10","y12", "y14"),
                   ordered=TRUE))) %>%
  pivot_wider(names_from = group, 
              values_from = key) %>% 
  rename(pds=p) %>%
  rename(space=s) 
a_long$age<-factor(a_long$age,levels=c( "y9", "y10","y12", "y14"),labels=c( "9", "10","12", "14"),ordered=TRUE)
a_long$age<-as.numeric(a_long$age)
```

```{r}
#correlations
pairs.panels(a[, 6:13])
```
```{r, include=FALSE}
# get the melt data
a_melt <- a %>% 
  pivot_longer(cols=y9space:y14pds, 
               names_to=c("age","value"), 
               names_pattern = "(y[0-9]{1,2})(\\w)",
               values_to="key",
               names_ptype=list(
                 age=factor(
                   levels=c("y9", "y10","y12", "y14"),
                   ordered=TRUE))) 
a_melt$age_baseline<-as.numeric(a_melt$age)
a_melt$age_baseline[a_melt$age_baseline == 1] <- 0
a_melt$age_baseline[a_melt$age_baseline == 2] <- 1
a_melt$age_baseline[a_melt$age_baseline == 4] <- 5
a_melt$value<-factor(a_melt$value,levels=c( "s", "p"),labels=c( "spatial", "puberty"),ordered=TRUE)
str(a_melt)

#re-order for convenience
a_melt <- a_melt[order(a_melt$id, a_melt$value), ]
```
```{r}
a_melt$puberty_on <- ifelse(a_melt$value == "puberty", 1, 0)
a_melt$spatial_on <- ifelse(a_melt$value == "spatial", 1, 0)

head(a_melt)
```

```{r}
apim_full <- lme(fixed= key ~  -1 +              #turns OFF intercept 
                                 puberty_on +           #pds intercept
                                 puberty_on:age_baseline +
                                 spatial_on +        # spatial ability intercept
                                 spatial_on:age_baseline,
                random = ~ -1 +puberty_on + spatial_on| id, 
               weights=varIdent(form = ~1 | puberty_on), # this invokes separate sigma^{2}_{e} for each gender
                  corr=corAR1(form = ~1 | id/sex/age_baseline), 
               data = a_melt,
               control=list(maxIter=10000),
               na.action = na.exclude)

summary(apim_full)
# not significant
```

```{r}
# APIM for males
a_melt_m<-a_melt %>% filter(sex==1)
apim_full_m <- lme(fixed= key ~  -1 +              #turns OFF intercept 
                                 puberty_on +           #pds intercept
                                 puberty_on:age_baseline +
                                 spatial_on +        # spatial ability intercept
                                 spatial_on:age_baseline,
                random = ~ -1 +puberty_on + spatial_on| id, 
               weights=varIdent(form = ~1 | puberty_on), # this invokes separate sigma^{2}_{e} for each gender
                  corr=corAR1(form = ~1 | id/age_baseline), 
               data = a_melt_m,
               control=list(maxIter=10000),
               na.action = na.exclude)

summary(apim_full_m)
# not significant
```

```{r}
# APIM for females
a_melt_f<-a_melt %>% filter(sex==0)
apim_full_f <- lme(fixed= key ~  -1 +              #turns OFF intercept 
                                 puberty_on +           #pds intercept
                                 puberty_on:age_baseline +
                                 spatial_on +        # spatial ability intercept
                                 spatial_on:age_baseline,
                random = ~ -1 +puberty_on + spatial_on| id, 
               weights=varIdent(form = ~1 | puberty_on), # this invokes separate sigma^{2}_{e} for each gender
                  corr=corAR1(form = ~1 | id/age_baseline), 
               data = a_melt_f,
               control=list(maxIter=10000),
               na.action = na.exclude)

summary(apim_full_f)
# not significant
```